Dependently Typed Languages in Statix

### 2022-10-12 paper-evcs/oasics-v2021-sample-article.tex
More work on aosics papers

@@

\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
\begin{CCSXML}
    <ccs2012>
    <concept>
    <concept_id>10011007.10011006.10011039.10011311</concept_id>
    <concept_desc>Software and its engineering~Semantics</concept_desc>
    <concept_significance>500</concept_significance>
    </concept>
    <concept>
    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
    <concept_desc>Software and its engineering~Functional languages</concept_desc>
    <concept_significance>300</concept_significance>
    </concept>
    <concept>
    <concept_id>10011007.10011006.10011041</concept_id>
    <concept_desc>Software and its engineering~Compilers</concept_desc>
    <concept_significance>300</concept_significance>
    </concept>
    </ccs2012>
\end{CCSXML}

@@

Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix.

Dependently typed languages are different from other languages because they allow types to be parameterized by values. This allows more rigorous reasoning over types and the values that are inhabited by a type. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

@@

The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or substitution respectively. To query the scope graph, use \verb|scopeGetName| or \verb|scopeGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found. Finally, \verb|empty_scope| returns a fresh empty scope.

\subsection{}

>>>>

% Added because the included oasics-logo-bw.pdf is pdf 1.6 and this was giving a warning
% Feel free to remove this if it breaks anything
\pdfminorversion=6

@@

\authorrunning{J. Brouwer, J. Cockx and A. Zwaan}

\Copyright{Jonathan Brouwer, Jesper Cockx and Aron Zwaan} 
\begin{CCSXML}
    <ccs2012>
    <concept>
    <concept_id>10011007.10011006.10011039.10011311</concept_id>
    <concept_desc>Software and its engineering~Semantics</concept_desc>
    <concept_significance>500</concept_significance>
    </concept>
    <concept>
    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
    <concept_desc>Software and its engineering~Functional languages</concept_desc>
    <concept_significance>300</concept_significance>
    </concept>
    <concept>
    <concept_id>10011007.10011006.10011041</concept_id>
    <concept_desc>Software and its engineering~Compilers</concept_desc>
    <concept_significance>300</concept_significance>
    </concept>
    </ccs2012>
\end{CCSXML}

@@

Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix.

Dependently typed languages are different from other languages, because they allow types to be parameterized by values. This allows more rigorous reasoning over types and the values that are inhabited by a type. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

@@

The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively. To query the scope graph, use \verb|scopeGetName| or \verb|scopeGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found. Finally, \verb|empty_scope| returns a fresh empty scope.

\subsection{Beta Reductions}

A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare. Beta reduction is done using Krivine abstract machines\cite{krivine}. We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications. The scope acts as the environment from \cite{krivine} paper, using \verb|NameSubst| to store substitutions. Furthermore, we can define \verb|expectBetaEq|, which asserts that two terms are equal. This calls \verb|expectBetaEq_| with \verb|betaReduceHead| applied.

\begin{lstlisting}
betaReduceHead : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
expectBetaEq   : (scope * Expr) * (scope * Expr)
expectBetaEq(e1, e2) :- 
    expectBetaEq_(betaReduceHead(e1, []), betaReduceHead(e2, [])).
expectBetaEq_  : (scope * Expr) * (scope * Expr)
\end{lstlisting}

\subsection{Type-checking programs}

We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
